\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-24 NAR_DRAFT.tex
correct, no true negatives
add space

\batcave divides the data into two classes: high- and low-confidence variants.
The high-confidence variants are used to estimate the mutational prior, which is then used to improve the calling of low-confidence variants.
Statistically, this is an empirical Bayesian approach~\citep{Robbins1954}, in which the high and low-confidence variants are treated as parallel experiments\citep{Morris1983,Efron2014}. 
In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
A key implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\citep{Rubanova2018a}, although those differences are relatively small.
A potential extension of the \batcave algorithm is process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
This approach might increasing sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.

@@

We evaluated \batcave with simulated tumors with three different mutation profiles and two real tumors.
The simulated diffuse and intermediate profiles (Fig.~\ref{NAR-simsigfig}A\&B) represent baseline profiles of lung and breast tumors, respectively, which the concentrated profile (Fig.~\ref{NAR-simsigfig}C) represents a tumor driven by a particular mutational process, such as UV exposure.
But mutational profiles are highly heterogeneous, so that concentrated profiles can be found in any tumor type (e.g., Fig.~\ref{NAR-realsigfig}C).
The two real data sets we considered are among the very few for which extensive validation of variant calls has been performed~\cite{Griffith2015, Shi2018}.
They happen, however, to have relatively flat mutation profiles (Fig.~\ref{NAR-realsigfig}A\&B), which reduces the expected advantage of \batcave over MuTect (Table~\ref{table:01}).
A more fundamental challenge of using these real data for testing callers is that only a subset of potential variants are validated.
This subset tends to be relatively easy to call, so both MuTect and \batcave have almost perfect precision and recall (Fig.~\ref{NAR-wes_fig} and Table~\ref{table:01}).
Moreover, few true negative sites are validated, so specificity is impossible to calculate. \rngcomment{Double-check my logic here...}

>>>>

\batcave divides the data into two classes: high- and low-confidence variants.
The high-confidence variants are used to estimate the mutational prior, which is then used to improve the calling of low-confidence variants.
Statistically, this is an empirical Bayesian approach~\citep{Robbins1954}, in which the high and low-confidence variants are treated as parallel experiments \citep{Morris1983,Efron2014}. 
In general, high-confidence variants tend to have relatively high allelic frequencies, and consequently tend to have arisen early in tumor development.
A key implicit assumption of our approach is that the mutational process does not change between high- and low-confidence variants, implying that the mutational profile of the tumor is temporally constant.
Recent studies have found differences in mutational profiles among variants of different allelic frequencies~\citep{Rubanova2018a}, although those differences are relatively small.
A potential extension of the \batcave algorithm is process potential variants in order of descending allelic frequency and to update the estimated mutational prior as the algorithm proceeds.
This approach might increasing sensitivity to low-frequency variants generated by recently-arisen mutational processes, at the cost of potentially increasing sensitivity to patterns of sequencing error.

@@

We evaluated \batcave with simulated tumors with three different mutation profiles and two real tumors.
The simulated diffuse and intermediate profiles (Fig.~\ref{NAR-simsigfig}A\&B) represent baseline profiles of lung and breast tumors, respectively, which the concentrated profile (Fig.~\ref{NAR-simsigfig}C) represents a tumor driven by a particular mutational process, such as UV exposure.
But mutational profiles are highly heterogeneous, so that concentrated profiles can be found in any tumor type (e.g., Fig.~\ref{NAR-realsigfig}C).
The two real data sets we considered are among the very few for which extensive validation of variant calls has been performed~\cite{Griffith2015, Shi2018}.
They happen, however, to have relatively flat mutation profiles (Fig.~\ref{NAR-realsigfig}A\&B), which reduces the expected advantage of \batcave over MuTect (Table~\ref{table:01}).
A more fundamental challenge of using these real data for testing callers is that only a subset of potential variants are validated.
This subset tends to be relatively easy to call, so both MuTect and \batcave have almost perfect precision and recall (Fig.~\ref{NAR-wes_fig} and Table~\ref{table:01}).
Moreover, few true negative sites are validated, so specificity is impossible to calculate.
